---
title: "Metal clustering"
author: "Bindoff, A."
output:
    html_document:
          code_folding: hide
---

`r Sys.time()`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load, message = F, warning = F}
library(readr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(latex2exp)
library(Rtsne)
library(vegan)

df <- read_csv("neuro_metal_nfl.csv", trim_ws = T)
names(df) <- gsub("\\b\\d+\\b", "", names(df)) # remove numbers
names(df) <- gsub("  ", "", names(df)) # remove extraneous blanks
names(df)[1:6] <- c("id", "age", "strain", "region", "sample", "dilution")
df$region[df$region == "Siatic nerve"] <- "Sciatic nerve"
df$id <- factor(df$id)
df$age <- factor(df$age)
df$strain <- gsub("[[:punct:]]", "", df$strain)
df$strain <- gsub("^WT AP.*", "WT_APPPS1", as.character(df$strain))
df$strain <- factor(df$strain)
df$region <- factor(df$region)

sel <- c(1,2,3,4,5,9,10,11,13,14,17,18,21,22,23,24)

df <- dplyr::select(df, sel) %>%# remove metals at or below detection limit
    filter(strain == "WT" | strain == "NFL", age == 5 | age == 22) %>% filter(id != 15) # nfl id == 15 was 19mo
df$strain <- factor(df$strain, levels = c("WT", "NFL"), labels = c("WT", "NFL-KO"))
df$age <- factor(df$age, levels = c(5, 22))
df <- na.omit(df)
df$Cu[df$Cu >= 320] <- NA 


`%nin%` <- Negate(`%in%`)

df <- dplyr::filter(df, id %nin% c("238", "3183", "161470"))

pal <- scico::scico(n = 6, palette = 'batlow')

```

## Heatmap

Metal concentrations are standardized across region and age using a centered log-ration transformation (Aitchison, J. 1986). Metals are ordinated by (statistical) similarity of composition. 

```{r echo = TRUE}
set.seed(42)

library(compositions)

B <- filter(df) %>%
  group_by(region, age) %>%
  mutate(Na = clr(Na),
         Mg = clr(Mg),
         Al = clr(Al),
         K =  clr(K), 
         Ca = clr(log(Ca)),
         Mn = clr(Mn),
         Fe = clr(Fe),
         Cu = clr(Cu),
         Zn = clr(Zn),
         Se = clr(Se),
         Rb = clr(Rb)) %>%
  ungroup() 

A <- select(B, Na, Mg, Al, K, Ca, Mn, Fe, Cu, Zn, Se, Rb)
#A[] <- apply(A, 2, function(x) ifelse(is.na(x), 0 , x))


# sort metals using Classical Multidimensional Scaling
D <- dist(A, method = "euclidean")
m3 <- cmdscale(D, k = 1)

tD <- dist(t(A), method = "euclidean")
m4 <- cmdscale(tD, k = 1)

# arrange samples and taxa by ordination above
B1 <- data.frame(B,
                 m3.Y = m3[,1]) %>%
  arrange(strain) %>%  # arrange samples
  mutate(sample = factor(sample, sample)) %>%
  reshape2::melt(id.vars = c("id", "age", "strain", "region", "sample", "m3.Y"))

mets <- tibble::rownames_to_column(data.frame(m4)) %>% arrange(m4)
B1$variable <- factor(B1$variable, mets$rowname)

# plot a heatmap
ggplot(B1, aes(x = variable, y = sample,
               fill = value)) +
  geom_tile() +
  scico::scale_fill_scico(palette = "vik", name = "Abundance (clr)") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6, colour = B1$strain)) +
  ylab("Sample (black WT, red NFL-KO)") +
  xlab("Metal")


metals <- mets$rowname
```


## Clustering

A stochastic neighbour embedding algorithm (tSNE) was fitted using data on metal composition for each sample. The algorithm was given no information on region, age, or genotype. Each datapoint represents a sample. Objects (datapoints) that are similar in high-dimensional space (the dimensions are the metals in this example) cluster together in low-dimensional space. There are no surprises here, the clusters clearly delineate into brain regions. I present this as a 'proof-of-concept' and to aid in understanding what tSNE can show us.  


```{r}
set.seed(321)
p <- df %>% group_by(region) %>%
  mutate(Al = scale(Al),
         Rb = scale(Rb),
         Zn = scale(Zn),
         Fe = scale(Fe),
         Se = scale(Se)) %>%
  ungroup() %>%
  select_split(Al, Zn, Fe, Rb, Se) %>%
  tsner(perplexity = 10, max_iter = 2000) %>%
  plotr() %>%
  order_taxa() %>%
  ggplot(aes(x = y1, y = y2, colour = strain, size = att, alpha = att, frame = taxa)) +
  geom_point() +
  theme_minimal()
  
plotly::ggplotly(p)




```


```{r}
set.seed(77)

A <- sapply(6:16, function(x) scale(df[, x]))
A[] <- apply(A, 2, function(x) ifelse(is.na(x), 0 , x))

m1 <- Rtsne(A, dim = 2, perplexity = 10, max_iter = 6000, theta = 0.1)

d <- data.frame(y1 = m1$Y[, 1], 
                y2 = m1$Y[, 2],
                strain = df$strain,
                id = df$id,
                region = df$region,
                age = df$age,
                strain = df$strain)

ggplot(d, aes(x = y1, y = y2, colour = region)) +
  geom_point(size = 1.5, alpha = 0.9) +
  geom_point(size = 4, alpha = 0.2) +
  scico::scale_color_scico_d(palette = "batlow", end = 0.9) +
  theme_minimal()

```

Clusters start to form around genotypes when the effects of region and age are partialled out. Size and transparency are varied according to standardized metal concentration. For a strong contrast, compare Rb with Fe by clicking on the interactive scale at the bottom of the plot. Clicking on `Play` will produce an animation. The metals have been ordered by similarity of composition within the sample, for example Zn and Ca covary more than Ca and Rb.  

```{r}

B <- filter(df) %>%
  group_by(region, age) %>%
  mutate(Na = scale(Na),
         Mg = scale(Mg),
         Al = scale(Al),
         K =  scale(K), 
         Ca = scale(log(Ca)),
         Mn = scale(Mn),
         Fe = scale(Fe),
         Cu = scale(Cu),
         Zn = scale(Zn),
         Se = scale(Se),
         Rb = scale(Rb)) %>%
  ungroup() %>%
  arrange(sample)

A <- select(B, Rb, Fe, Al)#metals)
A[] <- apply(A, 2, function(x) ifelse(is.na(x), 0 , x))

#A <- select(B, Al, Fe, Zn, Rb)


set.seed(888)
m1 <- Rtsne(A, dim = 2,
            perplexity = 7,
            max_iter = 10000, 
            theta = 0.05)

# make a data frame that assigns an attribute one 
# column/taxa at a time
foo <- function(x){
  g <- as.matrix(B[, x])
  B %>% mutate(att = g,
               Metal = x,
                    y1 = m1$Y[, 1],
                    y2 = m1$Y[, 2])
  
}

k <- lapply(metals, foo)
k <- dplyr::bind_rows(k)
k$Metal <- factor(k$Metal, rev(metals))

p1 <- ggplot(k, aes(x = y1, y = y2, colour = strain, size = att, alpha = att, frame = Metal, scale = 1.5, z = age)) +
  geom_point() +
  geom_point(size = .5, alpha = .8, show.legend = FALSE) +
  scale_colour_manual(values = pal[c(2, 4)]) +
  theme_minimal()

#plotly::plotly_json(p1)

plotly::ggplotly(p1)

```


```{r eval = FALSE}
set.seed(888)

# fit 10 models, pick the one with lowest KL divergence
ms <- lapply(1:10, function(x) Rtsne(A, dim = 2,
            perplexity = 7,
            max_iter = 10000, 
            theta = 0.05))

last_itercosts <- function(x){
  x$itercosts[length(x$itercosts)]
}

m1 <- ms[[which.min(lapply(ms, last_itercosts))]]

# make a data frame that assigns an attribute one 
# column/taxa at a time
foo <- function(x){
  g <- as.matrix(B[, x])
  B %>% mutate(att = g,
               Metal = x,
                    y1 = m1$Y[, 1],
                    y2 = m1$Y[, 2])
  
}

k <- lapply(metals, foo)
k <- dplyr::bind_rows(k)
k$Metal <- factor(k$Metal, rev(metals))

p1 <- ggplot(k, aes(x = y1, y = y2, colour = strain, size = att, alpha = att, frame = Metal, scale = 1.5)) +
  geom_point() +
  geom_point(size = .5, alpha = .8, show.legend = FALSE) +
  scale_colour_manual(values = pal[c(2, 4)]) +
  theme_minimal()

#plotly::plotly_json(p1)

plotly::ggplotly(p1)
```



```{r eval = FALSE}

B <- filter(df) %>%
  group_by(region, age) %>%
  mutate(Na = scale(as.vector(clr(Na))),
         Mg = scale(as.vector(clr(Mg))),
         Al = scale(as.vector(clr(Al))),
         K =  scale(as.vector(clr(K))), 
         Ca = scale(as.vector(clr(Ca))),
         Mn = scale(as.vector(clr(Mn))),
         Fe = scale(as.vector(clr(Fe))),
         Cu = scale(as.vector(clr(Cu))),
         Zn = scale(as.vector(clr(Zn))),
         Se = scale(as.vector(clr(Se))),
         Rb = scale(as.vector(clr(Rb)))) %>%
  ungroup()

A <- select(B, metals)
A[] <- apply(A, 2, function(x) ifelse(is.na(x), 0 , x))


set.seed(888)


ms <- lapply(1:10, function(x) Rtsne(A, dim = 2,
            perplexity = 7,
            max_iter = 10000, 
            theta = 0.05))
last_itercosts <- function(x){
  x$itercosts[length(x$itercosts)]
}

lapply(ms, last_itercosts)

m1 <- ms[[which.min(lapply(ms, last_itercosts))]]

# make a data frame that assigns an attribute one 
# column/taxa at a time
foo <- function(x){
  g <- as.matrix(B[, x])
  B %>% mutate(att = g,
               Metal = x,
                    y1 = m1$Y[, 1],
                    y2 = m1$Y[, 2])
  
}

k <- lapply(metals, foo)
k <- dplyr::bind_rows(k)
k$Metal <- factor(k$Metal, metals)

p1 <- ggplot(k, aes(x = y1, y = y2, colour = strain, size = att, alpha = att, frame = Metal, scale = 1.5)) +
  geom_point() +
  geom_point(size = .5, alpha = .8, show.legend = FALSE) +
  scale_colour_manual(values = pal[c(2, 4)]) +
  theme_minimal()

#plotly::plotly_json(p1)

plotly::ggplotly(p1)

```

## Hierarchical tSNE

I believe this to be a novel approach to hierarchical tSNE. A mixed model is used to estimate the population level effects, then conditional estimates are 'shrunk' towards the population mean using the random effects structure.  

```{r}
B <- df %>%
  group_by(region, age) %>%
  mutate(Na = scale(Na),
         Mg = scale(Mg),
         Al = scale(Al),
         K =  scale(K), 
         Ca = scale(Ca),
         Mn = scale(Mn),
         Fe = scale(Fe),
         Cu = scale(Cu),
         Zn = scale(Zn),
         Se = scale(Se),
         Rb = scale(Rb)) %>%
  ungroup() %>%
  arrange(sample)



tB <- reshape2::melt(B, id.vars = c("id", "age", "strain", "region", "sample"))
names(tB) <- c("id", "age", "strain", "region", "sample", "metal", "conc")
tB$conc[is.na(tB$conc)] <- 0
```

```{r}

lm1 <- glmmTMB::glmmTMB(conc ~ metal*region*age*strain + (metal|id/region), tB)

tB$fit <- predict(lm1, tB)

ggplot(tB, aes(x = conc, y = fit, colour = region)) +
  facet_wrap(~metal, scales = "free") +
  geom_abline(linetype = "dashed") +
  geom_point(alpha = 0.5) +
  scale_colour_viridis_d()

```

```{r eval = FALSE}
set.seed(1)
ggplot(tB, aes(x = region, y = fit, colour = strain)) +
  ggbeeswarm::geom_quasirandom(alpha = 0.6) +
  facet_wrap(~metal, scales = "free_y")
set.seed(1)
ggplot(tB, aes(x = region, y = conc, colour = strain)) +
  ggbeeswarm::geom_quasirandom(alpha = 0.6) +
  facet_wrap(~metal, scales = "free_y")

```

```{r, eval = FALSE}
A1 <- A
A <- select(tB, sample, metal, conc) %>% 
  tidyr::spread(metal, conc) %>%
  arrange(sample) %>%
  select(Rb, Se, K, Na, Cu, Fe, Mn, Al)
A1-A  ## should be zero!
```

```{r}

A <- select(tB, sample, metal, fit) %>% 
  tidyr::spread(metal, fit) %>%
  select(Rb, Se, K, Na, Cu, Fe, Mn, Al)

set.seed(888)
m1 <- Rtsne(A, dim = 2,
            perplexity = 7,
            max_iter = 12000, 
            theta = 0.05)

# make a data frame that assigns an attribute one 
# column/taxa at a time
foo <- function(x){
  g <- as.matrix(B[, x])
  B %>% mutate(att = g,
               Metal = x,
                    y1 = m1$Y[, 1],
                    y2 = m1$Y[, 2])
  
}

k <- lapply(metals, foo)
k <- dplyr::bind_rows(k)
k$Metal <- factor(k$Metal, rev(metals))

p1 <- ggplot(k, aes(x = y1, y = y2, colour = strain, size = att, alpha = att, frame = Metal, scale = 1.5)) +
  geom_point() +
  geom_point(size = .5, alpha = .8, show.legend = FALSE) +
  scale_colour_manual(values = pal[c(2, 4)]) +
  theme_minimal()

#plotly::plotly_json(p1)

plotly::ggplotly(p1)

```

## Ordination

Classical multi-dimensional (and non-metric, not shown) yield interpretable results, but the lack of clusters does not yield the same richness of information.  

```{r}
set.seed(888)
m1 <- cmdscale(dist(A), k = 2)

# make a data frame that assigns an attribute one 
# column/taxa at a time
foo <- function(x){
  g <- as.matrix(B[, x])
  B %>% mutate(att = g,
               Metal = x,
                    y1 = m1[, 1],
                    y2 = m1[, 2])
  
}

k <- lapply(metals, foo)
k <- dplyr::bind_rows(k)
k$Metal <- factor(k$Metal, rev(metals))

p1 <- ggplot(k, aes(x = y1, y = y2, colour = strain, size = att, alpha = att, frame = Metal, scale = 1.5)) +
  geom_point() +
  geom_point(size = .5, alpha = .8, show.legend = FALSE) +
  scale_colour_manual(values = pal[c(2, 4)]) +
  theme_minimal()

#plotly::plotly_json(p1)

plotly::ggplotly(p1)

```

## Going wide

Spreading the data in metal-brain-region columns may provide further richness


