---
title: "neuro_metal_nfl"
author: "Bindoff, A."
date: "24 May 2017"
output: html_document
---

### Executive Summary

Data from tissue metal analysis provided by Florey Institute were analysed for between group differences in NFL vs WT mice, 5mo vs 22mo, and different brain regions.

Multivariate ANOVA revealed significant differences in region, strain, age, and all two-way interactions. The dependent variables were log(Fe), log(Al), K, and Rb. A statistical appendix is provided below.

#### To do:

- outlier removal
- discuss the theoretical model to inform hypothesis testing
- deal with the multicollinearity (PCA?)
- if the theoretical model is uncertain, stepdown analysis? LDA?

### Statistical Appendix

#### Tissue metal analysis

NFL vs WT mice  
5mo v 22mo  
Brain regions  

Data are first loaded from a .csv file which was prepared using data returned from Florey Institute (Adlard). The following metals were detected at or below detection limits and are excluded from this analysis: Li, Cr,  Co, Ni, Sr, Ru,  Mo, Cd, Ba & Pt.   



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load, message = F, warning = F}
library(readr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(GGally)
df <- read_csv("neuro_metal_nfl.csv", trim_ws = T)
names(df) <- gsub("\\b\\d+\\b", "", names(df)) # remove numbers
names(df) <- gsub("  ", "", names(df)) # remove extraneous blanks
names(df)[1:6] <- c("id", "age", "strain", "region", "sample", "dilution")

df$id <- as.factor(df$id)
df$age <- as.factor(df$age)
df$strain <- gsub("[[:punct:]]", "", df$strain)
df$strain <- gsub("^WT AP.*", "WT_APPPS1", as.character(df$strain))
df$strain <- as.factor(df$strain)
df$region <- as.factor(df$region)

df <- select(df, c(1,2,3,4,5,9,10,11,13,14,17,18,21,22,23,24)) %>%# remove metals at or below detection limit
    filter(strain == "WT" | strain == "NFL", age == 5 | age == 22)
df <- na.omit(df)

```

  
Correlation matrix plots are produced for an informative visual exploration. From these it can be seen that metals accumulate differentially across brain regions (first plot). NFL strains may accumulate Fe, AL, K, & Rb differently to WT mice, so the next correlation matrix plot narrows down on these metals. An age x strain interaction appears to be present for Al and Rb.  


```{r exploratory plots, message = F, warning = F, eval = T}

ggpairs(df[,c(4,6:16)], aes(colour = region, alpha = 0.05), legend = c(1,1))

# df0 <- df[,c(2, 6:16)]
# ggpairs(df0, aes(colour = age, alpha = 0.05), legend = c(1,1))

# df0 <- df[,c(2, 3, 6:16)]
# ggpairs(df0, aes(colour = strain, alpha = 0.05), legend = c(1,1))

df0 <- df[, c(2, 3, 12, 8, 9, 16)]
ggpairs(df0, aes(colour = strain, alpha = 0.05), legend = c(1,1))


```
  
  
Plots are produced to assist in visualising age x strain x region relationships.

```{r exploratory plots2, message = F, warning = F, eval = T}


df0 <- melt(df[,c(1:4, 8, 11,12,13,14,16)], id.vars = c("id", "age", "strain", "region"))
df0$logvalue <- log(df0$value)
ggplot(df0, aes(x = strain, y = logvalue, colour = age, alpha = 0.05)) +
  geom_boxplot() +
  facet_grid(~variable, scales = "free")

df0 <- melt(df[,c(1:4, 7, 9, 10)], id.vars = c("id", "age", "strain", "region"))
df0$logvalue <- log(df0$value)
ggplot(df0, aes(x = strain, y = logvalue, colour = age, alpha = 0.05)) +
  geom_boxplot() +
  facet_grid(~variable, scales = "free")

df0 <- melt(df[,c(1:4, 8, 11,12,13,14,16)], id.vars = c("id", "age", "strain", "region"))
df0$logvalue <- log(df0$value)
ggplot(df0, aes(x = strain, y = logvalue, colour = variable, alpha = 0.05)) +
  geom_boxplot() +
  facet_wrap(~region, scales = "free", ncol = 3)

df0 <- melt(df[,c(1:4, 7, 9, 10)], id.vars = c("id", "age", "strain", "region"))
df0$logvalue <- log(df0$value)
ggplot(df0, aes(x = strain, y = logvalue, colour = variable, alpha = 0.05)) +
  geom_boxplot() +
  facet_wrap(~region, scales = "free", ncol = 3)

df0 <- melt(df[,c(1:4, 7:16)], id.vars = c("id", "age", "strain", "region"))
df0$logvalue <- log(df0$value)
ggplot(df0, aes(x = strain, y = logvalue, colour = age, alpha = 0.05)) +
  geom_boxplot() +
  facet_grid(~region, scales = "free")


```


Further distributional assumptions are checked.  Density plots suggest large departures from normality and the presence of outliers. Without *a priori* knowledge of these tests, it is not clear whether these outliers are measurement error (in which case they can be discarded), or reflect a normal biological range. A decision was made to retain outliers and investigate their effect on the models later. Density plots have been reproduced with log-transformed values, suggesting that transformation would be appropriate for most measures.

```{r histograms, message = F}
df0 <- melt(df[,c(1:4, 7:16)], id.vars = c("id", "age", "strain", "region"))
df0$logvalue <- log(df0$value)

ggplot(df0, aes(x = value)) +
  geom_density() +
  facet_wrap(~variable, scales = "free", ncol = 3)

ggplot(df0, aes(x = logvalue)) +
  geom_density() +
  facet_wrap(~variable, scales = "free", ncol = 3)


```

A series of factorial ANOVAs were fitted, and diagnostic plots for assumption checking. As expected, there were influential outliers and violated assumptions including normality (points on Q-Q plots departed from the dashed line) and homogeneity of variance (not equal variance across the residuals vs fitted plot).

```{r simple linear models}
par(mfrow = c(2,2))
anova(fit1 <- lm(Al ~ region*strain*age, df))
plot(fit1)
anova(fit2 <- lm(Rb ~ region*strain*age, df))
plot(fit2)
# anova(fit3 <- lm(Mg ~ region*strain*age, df))
anova(fit4 <- lm(K ~ region*strain*age, df))
plot(fit4)
anova(fit5 <- lm(Fe ~ region*strain*age, df))
plot(fit5)
anova(fit6 <- lm(Ca ~ region*strain*age, df))
plot(fit6)
anova(fit7 <- lm(Zn ~ region*strain*age, df))
plot(fit7)


```



We first log transform Al, K, & Fe values and remove observation 45 from the Al model and observation 184 from the Fe model and fit these models and diagnostic plots again. Al and Fe models were greatly improved, but K model suffered from the log transformation so the untransformed model was retained.

```{r remove outliers fit again}
par(mfrow = c(2,2))
anova(fit1 <- lm(log(Al) ~ region*strain*age, df[-45,]))
plot(fit1)
df$logAl <- log(df$Al)

anova(fit4 <- lm(log(K) ~ region*strain*age, df))
plot(fit4)

anova(fit5 <- lm(log(Fe) ~ region*strain*age, df[-184,]))
plot(fit5)
df$logFe <- log(df$Fe)

fit4 <- lm(K ~ region*strain*age, df[-c(167,124),])

anova(fit6 <- lm(log(Ca) ~ region*strain*age, df))
plot(fit6)
df$logCa <- log(df$Ca)

anova(fit7 <- lm(log(Zn) ~ region*strain*age, df))
plot(fit7)
df$logZn <- log(df$Zn)

```

Scale all variables to have mean = 0 and sd = 1, then run a MANOVA. Note that group sample sizes are not equal, but R makes the appropriate adjustments. No observations were excluded as outliers were not common among DVs and each had little effect on the model fit.

```{r MANOVA, message = F, echo = T}
df[,6:20] <- scale(df[,6:20])
fit1 <- manova(cbind(logAl, logFe, K, Rb, logCa, logZn) ~ region*strain*age, df)
summary(fit1)

```
  
There is no reason to model the three-way interaction, so this term is dropped, along with the DV, logCa.

```{r MANOVA2, message = F, echo = T}
fit2 <- manova(cbind(logAl, logFe, K, Rb, logZn) ~ 
                 region + strain + age +
                 region:strain +
                 region:age +
                 strain:age, df)
summary(fit2)
summary.aov(fit2)

```

