---
title: "neuro_metal_nfl_effect_sizes"
author: "Bindoff, A."
date: "9 June 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load, message = F, warning = F}
library(readr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(GGally)
library(sjPlot)
library(plotmo)
library(lme4)

df <- read_csv("neuro_metal_nfl.csv", trim_ws = T)
names(df) <- gsub("\\b\\d+\\b", "", names(df)) # remove numbers
names(df) <- gsub("  ", "", names(df)) # remove extraneous blanks
names(df)[1:6] <- c("id", "age", "strain", "region", "sample", "dilution")

df$id <- as.factor(df$id)
df$age <- as.factor(df$age)
df$strain <- gsub("[[:punct:]]", "", df$strain)
df$strain <- gsub("^WT AP.*", "WT_APPPS1", as.character(df$strain))
df$strain <- as.factor(df$strain)
df$region <- as.factor(df$region)

df <- select(df, c(1,2,3,4,5,9,10,11,13,14,17,18,21,22,23,24)) %>%# remove metals at or below detection limit
    filter(strain == "WT" | strain == "NFL", age == 5 | age == 22)
df <- na.omit(df)

```

 

```{r transformation, message = F, warning = F, eval = T}


df$logAl <- log(df$Al)
df$logFe <- log(df$Fe)
df$logCa <- log(df$Ca)
df$logZn <- log(df$Zn)
df[,6:20] <- scale(df[,6:20])

```


Some measures are log-transformed as per previous analysis, then all measures standardised to have $\mu = 0$ and $\sigma^2 = 1$


We are primarily interested in metals inside cells, and K is ubiquitous in cells, so it may provide a proxy for cell density in tissue samples. The correlation matrix plots show a strong correlation between K and Fe, K and Zn, and a very strong correlation between K and Rb. Conditioning on region, some of these correlations disappear (not shown explicitly). We know that K and Rb are exchangeable, and that the biological function of Rb is unknown (see introduction to Roberts et al. 2016 "Rubidium and potassium levels are altered in Alzheimer’s disease brain and blood but not in cerebrospinal fluid" for an overview).  This suggests that higher levels of Fe or Zn may be explained by a higher cell density in the tissue samples, particularly looking at different brain regions where we might expect differing cell densities.  



```{r correlationmatrixK, message = F, warning = F}
# fit2 <- manova(cbind(logAl, logFe, K, Rb, logZn) ~ 
df0 <- df[, c("region", "logAl", "logFe", "K", "Rb", "logZn")]
ggpairs(df0, aes(colour = region, alpha = 0.05), legend = c(1,1))



```

  
There are differences in K and Rb between NFL and wt mice, so we explore these differences more closely (looking at regions independently).  **lengthy output from t-tests has been suppressed**


```{r by_region}
ggplot(df, aes(x = K, y = logFe, colour = strain, alpha = 0.05)) + 
  geom_point() +
  geom_smooth(method = lm)

j_region <- c("Brain stem", "Cerebellum", "Cortex", "Hippocampus", "Olfactory bulb", "Siatic nerve")

Kj <- lapply(j_region, function(j) t.test(K ~ strain, data = filter(df, region == j)))
Rbj <- lapply(j_region, function(j) t.test(Rb ~ strain, data = filter(df, region == j)))
```


After a Bonferroni correction ($\alpha = .0083$), *no significant difference in K* between strains in different regions was observed, but *differences for Rb were observed for the Olfactory bulb*.

### age:strain interactions for Cortex region only

```{r linear_model_cortex}
df0 <- df
df <- filter(df, region == "Cortex")

summary(fit <- lm(logAl ~ strain*age, df))
plotmo(fit, method = "partdep")

summary(fit <- lm(logFe ~ strain*age, df))
plotmo(fit, method = "partdep")

summary(fit <- lm(K ~ strain*age, df))
plotmo(fit, method = "partdep")

summary(fit <- lm(Rb ~ strain*age, df))
plotmo(fit, method = "partdep")

summary(fit <- lm(logZn ~ strain*age, df))
plotmo(fit, method = "partdep")

```

#### including K as a covariate

In some sense this accounts for some of the small (but non-significant) differences in cell densities between strains. 

```{r K as covariate}
ggplot(df, aes(x = K, y = logFe, colour = strain, alpha = 0.05)) + 
  geom_point() +
  geom_smooth(method = lm)

summary(fit <- lm(logFe ~ K + strain*age, df))
plotmo(fit, method = "partdep")

```


### Region:strain, region:age, strain:age regression analysis 

```{r linear_model_region_interactions}
df <- df0
summary(fit <- lm(logAl ~ region + strain + age +
                 region:strain +
                 region:age +
                 strain:age, df))
plotmo(fit, method = "partdep")

summary(fit <- lm(logFe ~ region + strain + age +
                 region:strain +
                 region:age +
                 strain:age, df))
plotmo(fit, method = "partdep")

summary(fit <- lm(K ~ region + strain + age +
                 region:strain +
                 region:age +
                 strain:age, df))
plotmo(fit, method = "partdep")

summary(fit <- lm(Rb ~ region + strain + age +
                 region:strain +
                 region:age +
                 strain:age, df))
plotmo(fit, method = "partdep")

summary(fit <- lm(logZn ~ region + strain + age +
                 region:strain +
                 region:age +
                 strain:age, df))
plotmo(fit, method = "partdep")

```


  

