---
title: "CU, Zn, Fe in brains of NFL knock-out mice"
author: "Bindoff, A."
date: "27 July 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load, message = F, warning = F}
library(readr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(sjPlot)
library(plotmo)
library(lme4)
library(MuMIn)
library(lsr)
library(latex2exp)
library(lsmeans)
library(lmerTest)
mciMean <- function(x){
  return(setNames(c(round(mean(x), 3), round(ciMean(x), 3)), c("mean", "lwr", "upr")))
}


df <- read_csv("neuro_metal_nfl.csv", trim_ws = T)
names(df) <- gsub("\\b\\d+\\b", "", names(df)) # remove numbers
names(df) <- gsub("  ", "", names(df)) # remove extraneous blanks
names(df)[1:6] <- c("id", "age", "strain", "region", "sample", "dilution")
df$region[df$region == "Siatic nerve"] <- "Sciatic nerve"
df$id <- as.factor(df$id)
df$age <- as.factor(df$age)
df$strain <- gsub("[[:punct:]]", "", df$strain)
df$strain <- gsub("^WT AP.*", "WT_APPPS1", as.character(df$strain))
df$strain <- as.factor(df$strain)
df$region <- as.factor(df$region)

df <- dplyr::select(df, c(1,2,3,4,5,9,10,11,13,14,17,18,21,22,23,24)) %>%# remove metals at or below detection limit
    filter(strain == "WT" | strain == "NFL", age == 5 | age == 22)
df$strain <- factor(df$strain, levels = c("WT", "NFL"))
df$age <- factor(df$age, levels = c(5, 22))
df <- na.omit(df)

```

```{r transformation, message = F, warning = F, eval = T}


df$logAl <- log(df$Al)
df$logFe <- log(df$Fe)
df$logCa <- log(df$Ca)
df$logZn <- log(df$Zn)
df$logCu <- log(df$Cu)
# df[,6:ncol(df)] <- scale(df[,6:ncol(df)])

```
  
## Summary of findings

Tissue samples were taken from multiple sites (coded as `region`) in mouse brains as explained elsewhere. To account for individual effects and dependency between observations, a mixed effects model with mouse `id` fitted as a random intercept was identified by robust model selection procedures.

From this model specification, confidence intervals which take into account model uncertainty were computed. These results are presented in plots and tables, along with the results of ANOVA tests in the statistical appendix below.

There were significant `strain`:`age` interactions for measures of Zn & Fe, and significant `region`:`strain` interactions and main effect of `strain` for measures of Zn, Fe, & Cu. We expect main effects of `region` and `age` (and their interactions), and these are included for statistical adjustment only.

It is suggested that the reader review the interaction plots and ANOVA tables, ignoring the remaining output which is presented for completeness and future reference.


## Statistical appendix

### Exploratory plot

Violin plots show the distribution of each metal by age and strain. Bars are placed at mean and 95% CIs.

```{r exploratory plots2, message = F, warning = F, eval = T}


df0 <- melt(df[,c(1:4, 18, 20, 21)], id.vars = c("id", "age", "strain", "region"))
ggplot(df0, aes(x = strain, y = value, colour = age)) +
   geom_violin(draw_quantiles = c(0.025, 0.5, 0.975)) +
   facet_grid(~variable, scales = "free") +
   theme(panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))

df0 <- df
```

### Random effect of individual mouse

Line plots show the effect of individual mouse. Some animals fairly consistently have higher levels of Zn or Fe, some have consistently lower. Because we cannot control the effect of individual experimentally, we call it a "random effect" and model it as a distribution, rather than a fixed effect. This reduces the unexplained variance in the model, resulting in a more powerful model.

```{r}
cbPalette <- rep(c("#E69F00", "#0072B2", "#D55E00", "#56B4E9", "#009E73", "#F0E442", "#CC79A7"),
                 20)

ggplot(df, aes(x = region, y = logZn, group = factor(id))) +
  geom_line() +
  aes(colour = factor(id)) +
  scale_colour_manual(values=cbPalette) + 
  facet_grid(~ age + strain) +
  guides(colour = FALSE) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.text.x = element_text(angle = 90),
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("logZn ($\\mu$g/g)"), x = "region")

ggplot(df, aes(x = region, y = logFe, group = factor(id))) +
  geom_line() +
  aes(colour = factor(id)) +
  scale_colour_manual(values=cbPalette) + 
  facet_grid(~ age + strain) +
  guides(colour = FALSE) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_text(angle = 90),
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("logFe ($\\mu$g/g)"), x = "region")

```
### Mixed effects model

Tissue samples were taken from multiple brain and CNS sites for each animal, so in order to account for this non-independence, mixed effects models were fitted using `logZn` as the response variable. A number of models were fitted investigating fixed and random effects. A final model was selected by likelihood ratio test and $\Delta AIC$ -  

\[\hat{y} = region + age + strain + region*strain + strain*age + region*age + (1|id)\]


```{r, echo = T}
df <- dplyr::select(df0, c(id, age, strain, region, logFe, logZn, logCu))
m_1 <- lmer(logZn ~ strain*age + (1|region) + (1|id), df, REML = F)
m_2 <- lmer(logZn ~ region + strain*age + (1|id), df, REML = F )
m_3 <- lmer(logZn ~ region*age + strain*age + (1|id), df, REML = F)
m_4 <- lmer(logZn ~ region*age*strain + (1|id), df, REML = F)
m_5 <- lmer(logZn ~ region*strain + strain*age + region*age + (1|id), df, REML = F)
m_6 <- lmer(logZn ~ region*strain + strain*age + (1|id), df, REML = F)


anova(m_3, m_4, m_5, m_6)

```


```{r}
fit.cu <- lmer(logCu ~ region*strain + strain*age + region*age + (1|id), df)
fit.zn <- lmer(logZn ~ region*strain + strain*age + region*age + (1|id), df)
fit.fe <- lmer(logFe ~ region*strain + strain*age + region*age + (1|id), df)
```
  
```{r bootstrapping, warning = F, messsage = F}
set.seed(123)
nsim <- 1000
### logZn

# df.zn <- data.frame(region = factor(rep(levels(df$region), 4)),
#                     strain = factor(c(rep("wt", 6), rep("tg", 6))),
#                     age = factor(c(rep(5, 12), rep(22, 12))))
bootfit <- bootMer(fit.zn, FUN=function(x) predict(x, df, re.form = NA),
                   nsim = nsim,
                   parallel = "multicore",
                   ncpus = 3L)
df.zn <- df
df.zn$lwr <- apply(bootfit$t, 2, quantile, 0.025)
df.zn$upr <- apply(bootfit$t, 2, quantile, 0.975)
df.zn$fit <- apply(bootfit$t, 2, mean)

# ci <- confint(fit.zn, method = "boot")
# df.zn$lwr 

zn <- ggplot(df.zn, aes(x = age, y= fit, colour = strain, group = strain)) +
  geom_line(position = position_dodge(width = 0.1), aes(linetype=strain), size=.6) +
     geom_point(position = position_dodge(width = 0.1), size=3) + 
     geom_errorbar(position = position_dodge(width = 0.1),
                   aes(ymax=upr, ymin=lwr), width=.1) +
  facet_grid(~ region) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("logZn ($\\mu$g/g)"), x = "age (months)")

### logCu
df.cu <- df
bootfit <- bootMer(fit.cu, FUN=function(x) predict(x, df, re.form = NA),
                   nsim = nsim,
                   parallel = "multicore",
                   ncpus = 3L)

df.cu$lwr <- apply(bootfit$t, 2, quantile, 0.025)
df.cu$upr <- apply(bootfit$t, 2, quantile, 0.975)
df.cu$fit <- apply(bootfit$t, 2, mean)

cu <- ggplot(df.cu, aes(x = age, y= fit, colour = strain, group = strain)) +
  geom_line(position = position_dodge(width = 0.1), aes(linetype=strain), size=.6) +
     geom_point(position = position_dodge(width = 0.1), size=3) + 
     geom_errorbar(position = position_dodge(width = 0.1),
                   aes(ymax=upr, ymin=lwr), width=.1) +
  facet_grid(~ region) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("logCu ($\\mu$g/g)"), x = "age (months)")


### logFe
df.fe <- df
bootfit <- bootMer(fit.fe, FUN=function(x) predict(x, df, re.form = NA),
                   nsim = nsim,
                   parallel = "multicore",
                   ncpus = 3L)


df.fe$lwr <- apply(bootfit$t, 2, quantile, 0.025)
df.fe$upr <- apply(bootfit$t, 2, quantile, 0.975)
df.fe$fit <- apply(bootfit$t, 2, mean)

fe <- ggplot(df.fe, aes(x = age, y= fit, colour = strain, group = strain)) +
  geom_line(position = position_dodge(width = 0.1), aes(linetype=strain), size=.6) +
     geom_point(position = position_dodge(width = 0.1), size=3) + 
     geom_errorbar(position = position_dodge(width = 0.1),
                   aes(ymax=upr, ymin=lwr), width=.1) +
  facet_grid(~ region) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("logFe ($\\mu$g/g)"), x = "age (months)")

# fixed effects estimates (not bootstrapped)

k <- aggregate(logFe ~ region*strain + strain*age + region*age, df, mean)
m <- lm(logFe ~ region*strain + strain*age + region*age, df)
df.k <- cbind(k[,1:3], predict(m, k, interval = "confidence"))

fe.raw <- ggplot(df.k, aes(x = age, y = fit, colour = strain, group = strain)) +
  geom_line(position =  position_dodge(width = 0.1), aes(linetype = strain), size = 0.6) +
  geom_point(position = position_dodge(width = 0.1), size = 3) +
  geom_errorbar(position = position_dodge(width = 0.1),
                   aes(ymax=lwr, ymin=upr), width=.1) +
  facet_grid(~ region) + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("FIXED EFFECTS ONLY MODEL logFe ($\\mu$g/g)"), x = "age (months)")
#lsmeans(fit.zn, pairwise ~ age | strain)
#lsmeans(fit.zn, pairwise ~ strain | age)


```

### Interaction plots

After fitting the mixed effects models, 95% confidence intervals were obtained by bootstrapping using `r nsim` iterations. This is a robust (but computationally expensive) method of obtaining confidence intervals which take into account the conditional modes of the random effects (Bates, Machler, Bolker, & Walker (2014) https://arxiv.org/pdf/1406.5823.pdf ,see p 27), and the mild heterogeneity of residuals that was observed. These confidence intervals take into account all the sources of variability in the data that we *can* take into account, without making assumptions about the distribution of the residuals based on the mean and standard deviation.

```{r}
cu
zn
fe
```


### Confidence intervals (obtained by bootstrapping)

Tables of 95% CIs are produced and the results of ANOVA. **Our interest lies with the `strain` interactions**, we expect `region` and `age` differences which are of no interest in our experiment and are included for statistical adjustment.

Zn (log$\mu g/gm$)

```{r}


k <- aggregate(cbind(fit, lwr, upr) ~ region*strain*age, df.zn, FUN = mean)
k[,4:6] <- round(k[,4:6], 2)      
print(k)
anova(fit.zn)
```
  
Cu (log$\mu g/gm$)

```{r}
k <- aggregate(cbind(fit, lwr, upr) ~ region*strain*age, df.cu, FUN = mean)
k[,4:6] <- round(k[,4:6], 2)      
print(k)
anova(fit.cu)
```

  
Fe (log$\mu g/gm$)

```{r}
k <- aggregate(cbind(fit, lwr, upr) ~ region*strain*age, df.fe, FUN = mean)
k[,4:6] <- round(k[,4:6], 2)      
print(k)
anova(fit.fe)
```

