---
title: "CU, Zn, Fe in brains of NFL knock-out mice"
author: "Bindoff, A."
date: "10th August 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = F)
```

```{r load, message = F, warning = F}
library(readr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(car)
library(lme4)
library(MuMIn)
library(latex2exp)
library(lsmeans)
library(lmerTest)
mciMean <- function(x){
  return(setNames(c(round(mean(x), 3), round(ciMean(x), 3)), c("mean", "lwr", "upr")))
}


df <- read_csv("neuro_metal_nfl.csv", trim_ws = T)
names(df) <- gsub("\\b\\d+\\b", "", names(df)) # remove numbers
names(df) <- gsub("  ", "", names(df)) # remove extraneous blanks
names(df)[1:6] <- c("id", "age", "strain", "region", "sample", "dilution")
df$region[df$region == "Siatic nerve"] <- "Sciatic nerve"
df$id <- factor(df$id)
df$age <- factor(df$age)
df$strain <- gsub("[[:punct:]]", "", df$strain)
df$strain <- gsub("^WT AP.*", "WT_APPPS1", as.character(df$strain))
df$strain <- factor(df$strain)
df$region <- factor(df$region)

df <- dplyr::select(df, c(1,2,3,4,5,9,10,11,13,14,17,18,21,22,23,24)) %>%# remove metals at or below detection limit
    filter(strain == "WT" | strain == "NFL", age == 5 | age == 22) %>% filter(id != 15) # nfl id == 15 was 19mo
df$strain <- factor(df$strain, levels = c("WT", "NFL"))
df$age <- factor(df$age, levels = c(5, 22))
df <- na.omit(df)

```

```{r transformation, message = F, warning = F, eval = T}
### outlier checking
# ii <- which(df$id %in% c(234, 162382))
# df$Label[ii] <- as.character(df$id[ii])
# ggplot(df, aes(x = region, y = Cu, colour = id, label = Label)) +
#   geom_point() + geom_text(hjust = -.1, vjust = .8) +guides(colour = F)



df$Cu[df$Cu >= 320] <- NA 
df$logAl <- log(df$Al)
df$logFe <- log(df$Fe)
df$logCa <- log(df$Ca)
df$logZn <- log(df$Zn)
df$logCu <- log(df$Cu)
# df$logCu[df$logCu > 5.76] <- NA
# df[,6:ncol(df)] <- scale(df[,6:ncol(df)])

```
  


```{r, echo = F}
`%nin%` <- Negate(`%in%`)

df <- dplyr::select(df0, c(id, age, strain, region, logFe, logZn, Cu)) %>%
  filter(id %nin% c("238", "3183", "161470"))
```


```{r}
fit.cu <- lmer(Cu ~ region*strain + strain*age + region*age + (1|id), df)
fit.zn <- lmer(logZn ~ region*strain + strain*age + region*age + (1|id), df)
fit.fe <- lmer(logFe ~ region*strain + strain*age + region*age + (1|id), df)
```
  
```{r bootstrapping, warning = F, messsage = F}
set.seed(123)
nsim <- 1000
### logZn

# df.zn <- data.frame(region = factor(rep(levels(df$region), 4)),
#                     strain = factor(c(rep("wt", 6), rep("tg", 6))),
#                     age = factor(c(rep(5, 12), rep(22, 12))))
bootfit <- bootMer(fit.zn, FUN=function(x) predict(x, df, re.form = NA),
                   nsim = nsim,
                   parallel = "multicore",
                   ncpus = 3L)
df.zn <- df
df.zn$lwr <- apply(bootfit$t, 2, quantile, 0.025)
df.zn$upr <- apply(bootfit$t, 2, quantile, 0.975)
df.zn$fit <- apply(bootfit$t, 2, mean)

# confirm that bootMer gives the same mean as the sample mean
# k <- group_by(df.zn, region, age, strain) %>% summarise(fit = mean(logZn, na.rm = T))
# df.zn <- left_join(df.zn, k)


# ci <- confint(fit.zn, method = "boot")
# df.zn$lwr 
pal <- c("dodgerblue1", "firebrick")
lab <- c("WT", "TG")

zn <- ggplot(df.zn, aes(x = age, y= fit, colour = strain, group = strain)) +
  scale_colour_manual(values = pal,
                        name = element_blank(),
                        labels = lab) +
  geom_line(position = position_dodge(width = 0.2), aes(linetype=strain), size=.6, show.legend = F) +
     geom_point(position = position_dodge(width = 0.2), size=3) + 
     geom_errorbar(position = position_dodge(width = 0.2),
                   aes(ymax=upr, ymin=lwr), width=.5) +
  geom_jitter(width = 0.3, aes(x = age, y = logZn), alpha = 1/4) +
  facet_grid(~ region) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("Zn log_{e}($\\mu$g/g)"), x = "Age (months)")

### logCu
df.cu <- df
bootfit <- bootMer(fit.cu, FUN=function(x) predict(x, df, re.form = NA),
                   nsim = nsim,
                   parallel = "multicore",
                   ncpus = 3L)

df.cu$lwr <- apply(bootfit$t, 2, quantile, 0.025)
df.cu$upr <- apply(bootfit$t, 2, quantile, 0.975)
df.cu$fit <- apply(bootfit$t, 2, mean)

cu <- ggplot(df.cu, aes(x = age, y= fit, colour = strain, group = strain)) +
  scale_colour_manual(values = pal,
                        name = element_blank(),
                        labels = lab) +
  geom_line(position = position_dodge(width = 0.2), aes(linetype=strain), size=.6, show.legend = F) +
     geom_point(position = position_dodge(width = 0.2), size=3) + 
     geom_errorbar(position = position_dodge(width = 0.2),
                   aes(ymax=upr, ymin=lwr), width=.5) +
  geom_jitter(width = 0.3, aes(x = age, y = Cu), alpha = 1/4) +
  facet_grid(~ region) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("Cu ($\\mu$g/g)"), x = "Age (months)")


### logFe
df.fe <- df
bootfit <- bootMer(fit.fe, FUN=function(x) predict(x, df, re.form = NA),
                   nsim = nsim,
                   parallel = "multicore",
                   ncpus = 3L)


df.fe$lwr <- apply(bootfit$t, 2, quantile, 0.025)
df.fe$upr <- apply(bootfit$t, 2, quantile, 0.975)
df.fe$fit <- apply(bootfit$t, 2, mean)

fe <- ggplot(df.fe, aes(x = age, y= fit, colour = strain, group = strain)) +
  scale_colour_manual(values = pal,
                        name = element_blank(),
                        labels = lab) +
  geom_line(position = position_dodge(width = 0.2), aes(linetype=strain), size=.6, show.legend = F) +
     geom_point(position = position_dodge(width = 0.2), size=3) + 
     geom_errorbar(position = position_dodge(width = 0.2),
                   aes(ymax=upr, ymin=lwr), width=.5) +
  geom_jitter(width = 0.3, aes(x = age, y = logFe), alpha = 1/4) +
  facet_grid(~ region) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("Fe log_{e}($\\mu$g/g)"), x = "Age (months)")

# fixed effects estimates (not bootstrapped)

k <- aggregate(logFe ~ region*strain + strain*age + region*age, df, mean)
m <- lm(logFe ~ region*strain + strain*age + region*age, df)
df.k <- cbind(k[,1:3], predict(m, k, interval = "confidence"))

fe.raw <- ggplot(df.k, aes(x = age, y = fit, colour = strain, group = strain)) +
  geom_line(position =  position_dodge(width = 0.1), aes(linetype = strain), size = 0.6) +
  geom_point(position = position_dodge(width = 0.1), size = 3) +
  geom_errorbar(position = position_dodge(width = 0.1),
                   aes(ymax=lwr, ymin=upr), width=.1) +
  facet_grid(~ region) + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("FIXED EFFECTS ONLY MODEL logFe ($\\mu$g/g)"), x = "age (months)")

# calculate ICC for each model

a <- data.frame(VarCorr(fit.fe))$vcov
icc.fe <- a[1]/(a[1]+a[2])

a <- data.frame(VarCorr(fit.zn))$vcov
icc.zn <- a[1]/(a[1]+a[2])

a <- data.frame(VarCorr(fit.cu))$vcov
icc.cu <- a[1]/(a[1]+a[2])



#lsmeans(fit.zn, pairwise ~ age | strain)
#lsmeans(fit.zn, pairwise ~ strain | age)


```
## Results

To correct skewness, Zn and Fe variables were $log_e$-transformed prior to analysis. Mixed effects models were fitted using the `lme4` package in R. Models included a random intercept for each subject to account for non-independence between brain regions within animals. Type III F statistics were computed using Kenward-Roger approximation, and 95% confidence intervals were obtained by parametric bootstrapping.

Strain x age and strain x region interactions were investigated for each metal, in a model which accounted for region x age interactions. For Zn and Fe, there were significant strain x age (Zn: F(1, 27) = 9, *p* = .005; Fe: F(1, 27) = 11, *p* = .002) and strain x region (Zn: F(5, 140) = 8, *p* < .001; Fe: F(5, 140) = 3, *p* = .021) interactions, but only the strain x region interaction was significant for Cu (F(5, 139) = 4, *p* = .002). Figure __ shows the strength and direction of these interactions, with 95% CIs (unadjusted for multiple comparisons).




logZn ($\mu g/gm$)

```{r}


k <- aggregate(cbind(fit, lwr, upr) ~ region*strain*age, df.zn, FUN = mean)
k[,4:6] <- round(k[,4:6], 2)      
print(k)
print(anova(fit.zn, ddf = "Kenward-Roger"), digits = 2)
```
  
Cu ($\mu g/gm$)

```{r}
k <- aggregate(cbind(fit, lwr, upr) ~ region*strain*age, df.cu, FUN = mean)
k[,4:6] <- round(k[,4:6], 2)      
print(k)
print(anova(fit.cu, ddf = "Kenward-Roger"), digits = 2)
```

  
logFe ($\mu g/gm$)

```{r}
k <- aggregate(cbind(fit, lwr, upr) ~ region*strain*age, df.fe, FUN = mean)
k[,4:6] <- round(k[,4:6], 2)      
print(k)
print(anova(fit.fe, ddf = "Kenward-Roger"), digits = 2)
```



```{r}
fit.cu <- lmer(Cu ~ region*strain + strain*age + region*age + (1|id), df)
fit.zn.region.strain <- lmer(logZn ~ region*strain + strain*age + region*age + (1|id), df)
fit.fe <- lmer(logFe ~ region*strain + strain*age + region*age + (1|id), df)

f2 <- function(AB, A){
  (AB-A)/(1-AB)
}

## Zn

fit.zn.region.strain <- lmer(logZn ~ strain*age + region*age + (1|id), df)
fit.zn.strain.age <- lmer(logZn ~ region*strain + region*age + (1|id), df)

rABzn <- r.squaredGLMM(fit.zn)
rA <- r.squaredGLMM(fit.zn.region.strain)
marginal.region.strain.zn <- round(f2(rABzn[1], rA[1]),4)

rA <- r.squaredGLMM(fit.zn.strain.age)
marginal.strain.age.zn <- round(f2(rABzn[1], rA[1]), 4)

## Cu

fit.cu.region.strain <- lmer(Cu ~ strain*age + region*age + (1|id), df)
fit.cu.strain.age <- lmer(Cu ~ region*strain + region*age + (1|id), df)

rABcu <- r.squaredGLMM(fit.cu)
rA <- r.squaredGLMM(fit.cu.region.strain)
marginal.region.strain.cu <- round(f2(rABcu[1], rA[1]),4)

rA <- r.squaredGLMM(fit.cu.strain.age)
marginal.strain.age.cu <- round(f2(rABcu[1], rA[1]),4)

## Fe

fit.fe.region.strain <- lmer(logFe ~ strain*age + region*age + (1|id), df)
fit.fe.strain.age <- lmer(logFe ~ region*strain + region*age + (1|id), df)

rABfe <- r.squaredGLMM(fit.fe)
rA <- r.squaredGLMM(fit.fe.region.strain)
marginal.region.strain.fe <- round(f2(rABfe[1], rA[1]),4)

rA <- r.squaredGLMM(fit.fe.strain.age)
marginal.strain.age.fe <- round(f2(rABfe[1], rA[1]),4)


```

