---
title: "CU, Zn, Fe in brains of NFL knock-out mice"
author: "Bindoff, A."
date: "10th August 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = F)
```

```{r load, message = F, warning = F}
library(readr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(car)
library(lme4)
library(lmerTest)
library(MuMIn)
library(latex2exp)
library(lsmeans)
mciMean <- function(x){
  return(setNames(c(round(mean(x), 3), round(ciMean(x), 3)), c("mean", "lwr", "upr")))
}


df <- read_csv("neuro_metal_nfl.csv", trim_ws = T)
names(df) <- gsub("\\b\\d+\\b", "", names(df)) # remove numbers
names(df) <- gsub("  ", "", names(df)) # remove extraneous blanks
names(df)[1:6] <- c("id", "age", "strain", "region", "sample", "dilution")
df$region[df$region == "Siatic nerve"] <- "Sciatic nerve"
df$id <- factor(df$id)
df$age <- factor(df$age)
df$strain <- gsub("[[:punct:]]", "", df$strain)
df$strain <- gsub("^WT AP.*", "WT_APPPS1", as.character(df$strain))
df$strain <- factor(df$strain)
df$region <- factor(df$region)

df <- dplyr::select(df, c(1,2,3,4,5,9,10,11,13,14,17,18,21,22,23,24)) %>%# remove metals at or below detection limit
    filter(strain == "WT" | strain == "NFL", age == 5 | age == 22) %>% filter(id != 15) # nfl id == 15 was 19mo
df$strain <- factor(df$strain, levels = c("WT", "NFL"))
df$age <- factor(df$age, levels = c(5, 22))
df <- na.omit(df)

```

```{r transformation, message = F, warning = F, eval = T}
### outlier checking
# ii <- which(df$id %in% c(234, 162382))
# df$Label[ii] <- as.character(df$id[ii])
# ggplot(df, aes(x = region, y = Cu, colour = id, label = Label)) +
#   geom_point() + geom_text(hjust = -.1, vjust = .8) +guides(colour = F)



df$Cu[df$Cu >= 320] <- NA 
df$logAl <- log(df$Al)
df$logFe <- log(df$Fe)
df$logCa <- log(df$Ca)
df$logZn <- log(df$Zn)
df$logCu <- log(df$Cu)
# df$logCu[df$logCu > 5.76] <- NA
# df[,6:ncol(df)] <- scale(df[,6:ncol(df)])

```
  
## Summary of findings

Tissue samples were taken from multiple sites (coded as `region`) in mouse brains as explained elsewhere. To account for individual effects and dependency between observations, a mixed effects model with mouse `id` fitted as a random intercept was identified by robust model selection procedures.

From this model specification, confidence intervals which take into account model uncertainty were computed. These results are presented in plots and tables, along with the results of ANOVA tests in the statistical appendix below.

There were significant `strain`:`age` interactions for measures of Zn & Fe, and significant `region`:`strain` interactions and main effect of `strain` for measures of Zn, Fe, & Cu. We expect main effects of `region` and `age` (and their interactions), and these are included for statistical adjustment only.

The proportion of variance explained by the random effect of animal was 8.5% for log(Zn), 6% for log(Fe), and 10.4% for Cu. The standardized effect sizes, reported as Cohen's $f^2$, for strain x region and strain x age interactions was less than .08 (small), with the exception of the the strain x region interaction for log(Zn), which had an effect size $f^2 = .192$. The proportion of variance explained by the full models was 0.8, 0.71, and 0.78 for log(Fe), log(Zn) and Cu respectively, suggesting that most of the variance in the sample can be explained by age, region, and strain and their two-way interactions.

It is suggested that the reader review the interaction plots and ANOVA tables, ignoring the remaining output which is presented for completeness and future reference.


## Statistical appendix

### Sample characteristics

Due to _____ some data for individual mice were missing within regions. The number of mice observed for each condition is cross-tabulated below.

```{r}
xtabs(~strain + age + region, df)

```

### Exploratory plot

Violin plots show the distribution of each metal by age and strain. Bars are placed at mean and 95% CIs.

A single extreme value was removed, being the logCu measurement for mouse id = 234 in the Olfactory Bulb. This was considered to be very likely to be measurement error.

```{r exploratory plots2, message = F, warning = F, eval = T}


df0 <- melt(df[,c(1:4, 18, 20, 21)], id.vars = c("id", "age", "strain", "region"))
ggplot(df0, aes(x = strain, y = value, colour = age)) +
   geom_violin(draw_quantiles = c(0.025, 0.5, 0.975)) +
   # geom_boxplot() +
   facet_grid(~variable, scales = "free") +
   theme(panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))
df0 <- melt(df[,c(1:4, 13)], id.vars = c("id", "age", "strain", "region"))
ggplot(df0, aes(x = strain, y = value, colour = age)) +
   geom_violin(draw_quantiles = c(0.025, 0.5, 0.975)) +
   # geom_boxplot() +
   facet_grid(~variable, scales = "free") +
   theme(panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))
df0 <- df
```

### Random effect of individual mouse

Line plots show the effect of individual mouse. Some animals fairly consistently have higher levels of Zn or Fe, some have consistently lower. Because we cannot control the effect of individual experimentally, we call it a "random effect" and model it as a distribution, rather than a fixed effect. This reduces the unexplained variance in the model, resulting in a more powerful model.  *This group level variance is summarised in the ICC statistic for the full (conditonal) model, detailed below.*  


```{r}
cbPalette <- rep(c("#E69F00", "#0072B2", "#D55E00", "#56B4E9", "#009E73", "#F0E442", "#CC79A7"), 20)

ggplot(df, aes(x = region, y = logZn, group = factor(id))) +
  geom_line() +
  aes(colour = factor(id)) +
  scale_colour_manual(values=cbPalette) + 
  facet_grid(~ age + strain) +
  # guides(colour = FALSE) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.text.x = element_text(angle = 90),
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("logZn ($\\mu$g/g)"), x = "region")

ggplot(df, aes(x = region, y = logFe, group = factor(id))) +
  geom_line() +
  aes(colour = factor(id)) +
  scale_colour_manual(values=cbPalette) + 
  facet_grid(~ age + strain) +
  # guides(colour = TRUE) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_text(angle = 90),
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("logFe ($\\mu$g/g)"), x = "region")

```
### Mixed effects model

Tissue samples were taken from multiple brain and CNS sites for each animal, so in order to account for this non-independence, mixed effects models were fitted using `logZn` as the response variable. A number of models were fitted investigating fixed and random effects. A final model was selected by likelihood ratio test and $\Delta AIC$ -  

\[y = region + age + strain + region*strain + strain*age + region*age + (1|id)\]


```{r, echo = T}
`%nin%` <- Negate(`%in%`)

df <- dplyr::select(df0, c(id, age, strain, region, logFe, logZn, Cu)) %>%
  filter(id %nin% c("238", "3183", "161470"))
m_1 <- lmer(logZn ~ strain*age + (1|region) + (1|id), df, REML = F)
m_2 <- lmer(logZn ~ region + strain*age + (1|id), df, REML = F )
m_3 <- lmer(logZn ~ region*age + strain*age + (1|id), df, REML = F)
m_4 <- lmer(logZn ~ region*age*strain + (1|id), df, REML = F)
m_5 <- lmer(logZn ~ region*strain + strain*age + region*age + (1|id), df, REML = F)
m_6 <- lmer(logZn ~ region*strain + strain*age + (1|id), df, REML = F)
m_7 <- lmer(logZn ~ region*strain + strain*age + region*age + (1|id) + (1|region), df, REML = F)

anova(m_3, m_4, m_5, m_6)

anova(m_5, m_7)

```


```{r}
fit.cu <- lmer(Cu ~ region*strain + strain*age + region*age + (1|id), df)
fit.zn <- lmer(logZn ~ region*strain + strain*age + region*age + (1|id), df)
fit.fe <- lmer(logFe ~ region*strain + strain*age + region*age + (1|id), df)
```
  
```{r bootstrapping, warning = F, messsage = F}
set.seed(123)
nsim <- 1000
### logZn

# df.zn <- data.frame(region = factor(rep(levels(df$region), 4)),
#                     strain = factor(c(rep("wt", 6), rep("tg", 6))),
#                     age = factor(c(rep(5, 12), rep(22, 12))))
bootfit <- bootMer(fit.zn, FUN=function(x) predict(x, df, re.form = NA),
                   nsim = nsim,
                   parallel = "multicore",
                   ncpus = 3L)
df.zn <- df
df.zn$lwr <- apply(bootfit$t, 2, quantile, 0.025)
df.zn$upr <- apply(bootfit$t, 2, quantile, 0.975)
df.zn$fit <- apply(bootfit$t, 2, mean)

# confirm that bootMer gives the same mean as the sample mean
# k <- group_by(df.zn, region, age, strain) %>% summarise(fit = mean(logZn, na.rm = T))
# df.zn <- left_join(df.zn, k)


# ci <- confint(fit.zn, method = "boot")
# df.zn$lwr 
pal <- c("dodgerblue1", "firebrick")
lab <- c("WT", "TG")

zn <- ggplot(df.zn, aes(x = age, y= fit, colour = strain, group = strain)) +
  scale_colour_manual(values = pal,
                        name = element_blank(),
                        labels = lab) +
  geom_line(position = position_dodge(width = 0.2), aes(linetype=strain), size=.6, show.legend = F) +
     geom_point(position = position_dodge(width = 0.2), size=3) + 
     geom_errorbar(position = position_dodge(width = 0.2),
                   aes(ymax=upr, ymin=lwr), width=.5) +
  geom_jitter(width = 0.3, aes(x = age, y = logZn), alpha = 1/4) +
  facet_grid(~ region) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("Zn log_{e}($\\mu$g/g)"), x = "Age (months)")

### logCu
df.cu <- df
bootfit <- bootMer(fit.cu, FUN=function(x) predict(x, df, re.form = NA),
                   nsim = nsim,
                   parallel = "multicore",
                   ncpus = 3L)

df.cu$lwr <- apply(bootfit$t, 2, quantile, 0.025)
df.cu$upr <- apply(bootfit$t, 2, quantile, 0.975)
df.cu$fit <- apply(bootfit$t, 2, mean)

cu <- ggplot(df.cu, aes(x = age, y= fit, colour = strain, group = strain)) +
  scale_colour_manual(values = pal,
                        name = element_blank(),
                        labels = lab) +
  geom_line(position = position_dodge(width = 0.2), aes(linetype=strain), size=.6, show.legend = F) +
     geom_point(position = position_dodge(width = 0.2), size=3) + 
     geom_errorbar(position = position_dodge(width = 0.2),
                   aes(ymax=upr, ymin=lwr), width=.5) +
  geom_jitter(width = 0.3, aes(x = age, y = Cu), alpha = 1/4) +
  facet_grid(~ region) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("Cu ($\\mu$g/g)"), x = "Age (months)")


### logFe
df.fe <- df
bootfit <- bootMer(fit.fe, FUN=function(x) predict(x, df, re.form = NA),
                   nsim = nsim,
                   parallel = "multicore",
                   ncpus = 3L)


df.fe$lwr <- apply(bootfit$t, 2, quantile, 0.025)
df.fe$upr <- apply(bootfit$t, 2, quantile, 0.975)
df.fe$fit <- apply(bootfit$t, 2, mean)

fe <- ggplot(df.fe, aes(x = age, y= fit, colour = strain, group = strain)) +
  scale_colour_manual(values = pal,
                        name = element_blank(),
                        labels = lab) +
  geom_line(position = position_dodge(width = 0.2), aes(linetype=strain), size=.6, show.legend = F) +
     geom_point(position = position_dodge(width = 0.2), size=3) + 
     geom_errorbar(position = position_dodge(width = 0.2),
                   aes(ymax=upr, ymin=lwr), width=.5) +
  geom_jitter(width = 0.3, aes(x = age, y = logFe), alpha = 1/4) +
  facet_grid(~ region) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("Fe log_{e}($\\mu$g/g)"), x = "Age (months)")

# fixed effects estimates (not bootstrapped)

k <- aggregate(logFe ~ region*strain + strain*age + region*age, df, mean)
m <- lm(logFe ~ region*strain + strain*age + region*age, df)
df.k <- cbind(k[,1:3], predict(m, k, interval = "confidence"))

fe.raw <- ggplot(df.k, aes(x = age, y = fit, colour = strain, group = strain)) +
  geom_line(position =  position_dodge(width = 0.1), aes(linetype = strain), size = 0.6) +
  geom_point(position = position_dodge(width = 0.1), size = 3) +
  geom_errorbar(position = position_dodge(width = 0.1),
                   aes(ymax=lwr, ymin=upr), width=.1) +
  facet_grid(~ region) + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(y = TeX("FIXED EFFECTS ONLY MODEL logFe ($\\mu$g/g)"), x = "age (months)")

# calculate ICC for each model

a <- data.frame(VarCorr(fit.fe))$vcov
icc.fe <- a[1]/(a[1]+a[2])

a <- data.frame(VarCorr(fit.zn))$vcov
icc.zn <- a[1]/(a[1]+a[2])

a <- data.frame(VarCorr(fit.cu))$vcov
icc.cu <- a[1]/(a[1]+a[2])



#lsmeans(fit.zn, pairwise ~ age | strain)
#lsmeans(fit.zn, pairwise ~ strain | age)


```

### Intra-class correlation

The proportion of variance explained by random effects for each (conditional) model is:
$ICC_{Zn} =$ `r icc.zn`  
$ICC_{Fe} =$ `r icc.fe`
$ICC_{Cu} =$ `r icc.cu`


### Interaction plots

After fitting the mixed effects models, 95% confidence intervals were obtained by bootstrapping using `r nsim` iterations. This is a robust (but computationally expensive) method of obtaining confidence intervals which take into account the conditional modes of the random effects (Bates, Machler, Bolker, & Walker (2014) https://arxiv.org/pdf/1406.5823.pdf ,see p 27), and the mild heterogeneity of residuals that was observed. These confidence intervals take into account all the sources of variability in the data that we *can* take into account, without making assumptions about the distribution of the residuals based on the mean and standard deviation.

```{r, warning = F}
cu
zn
fe
```


### Confidence intervals (obtained by bootstrapping)

Tables of 95% CIs are produced and the results of ANOVA. **Our interest lies with the `strain` interactions**, we expect `region` and `age` differences which are of no interest in our experiment and are included for statistical adjustment.

logZn ($\mu g/gm$)

```{r}


k <- aggregate(cbind(fit, lwr, upr) ~ region*strain*age, df.zn, FUN = mean)
k[,4:6] <- round(k[,4:6], 2)      
print(k)
print(anova(fit.zn, ddf = "Kenward-Roger"), digits = 2)
```

#### Sidak-adjusted comparisons, strain*age

Cortex  


```{r}
region.zn <- lm(logZn ~ strain*age, filter(df, region == "Cortex"))
region.zn.contr <- lsmeans(region.zn, ~strain*age)
contrast(region.zn.contr, alpha = .05, by = "age", adjust = NULL)
```
  
  Hippocampus  
  
```{r}
region.zn <- lm(logZn ~ strain*age, filter(df, region == "Hippocampus"))
region.zn.contr <- lsmeans(region.zn, ~strain *age)
contrast(region.zn.contr, alpha = .05, by = "age", adjust = NULL)
```
  
  Sciatic nerve  
  
```{r}
region.zn <- lm(logZn ~ strain*age, filter(df, region == "Sciatic nerve"))
region.zn.contr <- lsmeans(region.zn, ~strain*age)
contrast(region.zn.contr, alpha = .05, by = "age", adjust = NULL)

# strain.region <- lsmeans(fit.zn, ~strain*region)
# strain.age <- lsmeans(fit.zn, ~strain*age)
# contrast(strain.age, alpha = 0.05, method = "pairwise", adjust= "Tukey")
```


```{r, eval = T}

qqnorm(resid(fit.zn))
qqline(resid(fit.zn))
qqnorm(resid(fit.cu))
qqline(resid(fit.cu))
qqnorm(resid(fit.fe))
qqline(resid(fit.fe))
```
  
Cu ($\mu g/gm$)

```{r}
k <- aggregate(cbind(fit, lwr, upr) ~ region*strain*age, df.cu, FUN = mean)
k[,4:6] <- round(k[,4:6], 2)      
print(k)
print(anova(fit.cu, ddf = "Kenward-Roger"), digits = 2)
```
#### Sidak-adjusted comparisons, strain*age

Cortex  


```{r}
region.cu <- lm(Cu ~ strain*age, filter(df, region == "Cortex"))
region.cu.contr <- lsmeans(region.cu, ~strain*age)
contrast(region.cu.contr, alpha = .05, by = "age", adjust = NULL)
```
  
  Hippocampus  
  
```{r}
region.cu <- lm(Cu ~ strain*age, filter(df, region == "Hippocampus"))
region.cu.contr <- lsmeans(region.cu, ~strain*age)
contrast(region.cu.contr, alpha = .05, by = "age", adjust = NULL)
```
  
  Sciatic nerve  
  
```{r}
region.cu <- lm(Cu ~ strain*age, filter(df, region == "Sciatic nerve"))
region.cu.contr <- lsmeans(region.cu, ~strain*age)
contrast(region.cu.contr, alpha = .05, by = "age", adjust = NULL)
```
  
logFe ($\mu g/gm$)

```{r}
k <- aggregate(cbind(fit, lwr, upr) ~ region*strain*age, df.fe, FUN = mean)
k[,4:6] <- round(k[,4:6], 2)      
print(k)
print(anova(fit.fe, ddf = "Kenward-Roger"), digits = 2)
```

#### Sidak-adjusted comparisons, strain*age

Cortex  


```{r}
region.fe <- lm(logFe ~ strain*age, filter(df, region == "Cortex"))
region.fe.contr <- lsmeans(region.fe, ~strain*age)
contrast(region.fe.contr, alpha = .05, by = "age", adjust = NULL)
```
  
  Hippocampus  
  
```{r}
region.fe <- lm(logFe ~ strain*age, filter(df, region == "Hippocampus"))
region.fe.contr <- lsmeans(region.fe, ~strain*age)
contrast(region.fe.contr, alpha = .05, by = "age", adjust = NULL)
```
  
  Sciatic nerve  
  
```{r}
region.fe <- lm(logFe ~ strain*age, filter(df, region == "Sciatic nerve"))
region.fe.contr <- lsmeans(region.fe, ~strain*age)
contrast(region.fe.contr, alpha = .05, by = "age", adjust = NULL)
```

### Effect Sizes

We have already selected a parsimonious model using statistics that compare the relative fit of competing models, but as yet have no estimate of the absolute fit of the model (for each metal). Nakagawa & Schielzeth (2012) propose a general and simple method for obtaining $R^2$ from mixed models. We estimate the  *marginal* $R^2$ which describes the proportion of variance explained by the fixed factors,
\[R^2_{LMM(m)} = \frac{\sigma^2_f}{\sigma^2_f + \Sigma\sigma^2_l + \sigma^2_e}\]

Where $\sigma^2_f$ is the fixed-effects variance, $\sigma^2_l$ are the (summed) random variance components, and $\sigma^2_e$ is the residual variance. We also report the *conditional* $R^2$ which can be interpreted as the variance explained by the full mixed model (Nakagawa & Schielzeth, 2012).


```{r}
fit.cu <- lmer(Cu ~ region*strain + strain*age + region*age + (1|id), df)
fit.zn.region.strain <- lmer(logZn ~ region*strain + strain*age + region*age + (1|id), df)
fit.fe <- lmer(logFe ~ region*strain + strain*age + region*age + (1|id), df)

f2 <- function(AB, A){
  (AB-A)/(1-AB)
}

## Zn

fit.zn.region.strain <- lmer(logZn ~ strain*age + region*age + (1|id), df)
fit.zn.strain.age <- lmer(logZn ~ region*strain + region*age + (1|id), df)

rABzn <- r.squaredGLMM(fit.zn)
rA <- r.squaredGLMM(fit.zn.region.strain)
marginal.region.strain.zn <- round(f2(rABzn[1], rA[1]),4)

rA <- r.squaredGLMM(fit.zn.strain.age)
marginal.strain.age.zn <- round(f2(rABzn[1], rA[1]), 4)

## Cu

fit.cu.region.strain <- lmer(Cu ~ strain*age + region*age + (1|id), df)
fit.cu.strain.age <- lmer(Cu ~ region*strain + region*age + (1|id), df)

rABcu <- r.squaredGLMM(fit.cu)
rA <- r.squaredGLMM(fit.cu.region.strain)
marginal.region.strain.cu <- round(f2(rABcu[1], rA[1]),4)

rA <- r.squaredGLMM(fit.cu.strain.age)
marginal.strain.age.cu <- round(f2(rABcu[1], rA[1]),4)

## Fe

fit.fe.region.strain <- lmer(logFe ~ strain*age + region*age + (1|id), df)
fit.fe.strain.age <- lmer(logFe ~ region*strain + region*age + (1|id), df)

rABfe <- r.squaredGLMM(fit.fe)
rA <- r.squaredGLMM(fit.fe.region.strain)
marginal.region.strain.fe <- round(f2(rABfe[1], rA[1]),4)

rA <- r.squaredGLMM(fit.fe.strain.age)
marginal.strain.age.fe <- round(f2(rABfe[1], rA[1]),4)


```

$R^2_{logFe} =$ `r round(rABfe,4)` (marginal, conditional)  
$R^2_{logZn} =$ `r round(rABzn,4)`  
$R^2_{Cu} =$ `r round(rABcu,4)`


An estimate of *local effect size* is calculated for interaction terms using *Cohen's $f^2$* statistic,    
\[f^2 = \frac{R^2_{AB} - R^2_{A}}{1 - R^2_{AB}}\]  
Where $R^2_{AB}$ is the marginal co-efficient of determination estimated for the full model (as described above), and $R^2_{A}$ is the marginal co-efficient of determination estimated for the model dropping the interaction term of interest. Care must be taken to interpret these effect sizes in context, and they are provided as standardized measures which are of practical benefit in comparing results between studies.



#### logFe

Marginal effect of region:strain interaction:    `r marginal.region.strain.fe`  
Marginal effect of strain:age interaction:        `r marginal.strain.age.fe`  

#### logZn

Marginal effect of region:strain interaction:    `r marginal.region.strain.zn`  
Marginal effect of strain:age interaction:        `r marginal.strain.age.zn`  

#### logCu

Marginal effect of region:strain interaction:    `r marginal.region.strain.cu`  
Marginal effect of strain:age interaction:        `r marginal.strain.age.cu`  
